---
title: "Nonstationary Regional Frequency Analysis with Peaks-Over-Threshold Data"
author: "Martin Roth"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, fig.show='hold'}
library(data.table)
library(ggplot2)
admNL <- raster::getData('GADM', country='NL', path="../inst/extdata", level=0)
fadmNL <- fortify(admNL)
if(!file.exists("../inst/extdata/NL_eobs_rr_0.50.rds")) {
  data  <- eobsR::importEOBS('rr', '1950/2015', admNL, "0.50reg")
  saveRDS(data, file = "../inst/extdata/NL_eobs_rr_0.50.rds")
} else {
  data <- readRDS("../inst/extdata/NL_eobs_rr_0.50.rds")
}
data <- data[month %in% c(1,2,12),]
knitr::kable(head(data))
```

```{r, fig.show='hold'}
ggplot(fadmNL, aes(x = long, y = lat, group = group)) +
  geom_path() + coord_map() + 
  geom_tile(aes(x = lon, y = lat, fill = mRR, group = NULL), data = 
              data[, .(RR = max(rr)),
                   by = .(year, lat, lon)][, .(mRR = mean(RR)),
                                           by = .(lat, lon)],
            alpha = 0.5) + 
  scale_fill_distiller(palette = "Spectral", guide =guide_legend(reverse=TRUE))
```


- plot winter maxima
- use declustering
- reproduce figures 3 and 4


Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
